Tracing the Meta-Level: PyPy's Tracing JIT Compiler

### 2009-04-08 talk/icooolps2009/paper.tex
Check in my changes, mostly typos, and let Carl Friedrich deal with the extra page :-)

Let's look at how hints would need to be applied to the example interpreter
from Figure \ref{fig:tlr-basic}. Figure \ref{fig:tlr-full} shows the relevant
parts of the interpreter with hints applied. To apply hints one needs a
subclass of \texttt{JitDriver} that lists all the variables of the bytecode
loop. The variables are classified into two groups, red variables and green
variables. The green variables are those that the tracing JIT should consider to
be part of the program counter of the language interpreter. In the case of the
example, the \texttt{pc} variable is obviously part of the program counter.
However, the \texttt{bytecode} variable is also counted as green, since the
\texttt{pc} variable is meaningless without the knowledge of which bytecode
string is currently being interpreted. All other variables are red (the fact
that red variables need to be listed explicitly too is an implementation
detail).

@@

For the small example the hints look like a lot of work. However, the number of
hints is essentially constant no matter how large the interpreter is, which
makes it less significant for larger interpreters.

@@

The simple insight how to improve the situation is that most of the operations
in the trace are actually concerned with manipulating the bytecode string and
the program counter. Those are stored in variables that are ``green'' (\ie they
are part of the position key).  This means that the tracer checks that those
variables have some fixed value at the beginning of the loop (they may well
change over the course of the loop, though). In the example of Figure
\ref{fig:trace-no-green-folding} the check would be that at the beginning of the
trace the \texttt{pc} variable is \texttt{4} and the \texttt{bytecode} variable
is the bytecode string corresponding to the square function. Therefore it is
possible to constant-fold computations on them away,
as long as the operations are side-effect free. Since strings are immutable in
RPython, it is possible to constant-fold the \texttt{strgetitem} operation. The
\texttt{int\_add} are additions of the green variable \texttt{pc} and a constant
number, so they can be folded away as well.

@@

From then on things proceed like described in Section \ref{sect:tracing}. The
tracing interpreter tries to find a loop in the user program, if it finds one it
will produce machine code for that loop and this machine code will be
immediately executed. The machine code is executed until a guard fails. Then the
execution should fall back to normal interpretation by the language interpreter.
This falling back is possibly a complex process, since the guard failure can
have occurred arbitrarily deep in a helper function of the language interpreter,
which would make it hard to rebuild the state of the language interpreter and
let it run from that point (\eg this would involve building a potentially deep
C stack). Instead the falling back is achieved by a special \emph{fallback
interpreter} which runs the language interpreter and the user program from the
point of the guard failure. The fallback interpreter is essentially a variant of
the tracing interpreter that does not keep a trace. The fallback interpreter
runs until execution reaches a safe point where it is easy to let the C version
of the language interpreter resume its operation. Usually this means that the
fallback interpreter executes at most one bytecode operation of the language
interpreter. After the language interpreter takes over again, the whole process
starts again.

Machine code production is done via a well-defined interface to an assembler
backend. This makes it possible to easily port the tracing JIT to various
architectures (including, we hope, to virtual machines such as the JVM where
backend could generate bytecode at runtime). At the moment the only implemented
backend is a 32-bit Intel-x86 backend.

@@

The first round of benchmarks (Figure \ref{fig:bench1}) are timings of the
example interpreter given in Figure \ref{fig:tlr-basic} computing
the square of 10000000\footnote{the result will overflow, but for smaller numbers the
running time is not long enough to sensibly measure it}
using the bytecode of Figure \ref{fig:square}. The results for various
configurations are as follows:

@@

\textbf{Benchmark 3:} The tracing JIT is enabled and hints as in Figure
\ref{fig:tlr-full} are applied. This means that the interpreter loop is unrolled
so that it corresponds to the loop in the square function. However, constant folding of green
variables is disabled, therefore the resulting machine code corresponds to the
trace in Figure \ref{fig:trace-no-green-folding}. This by alone brings no
improvement over the previous case.

@@

\begin{figure}
\noindent
\begin{tabular}{|l|l|r|r|}
\hline
& &Time (ms) &speedup\\
\hline
1 &compiled to C, no JIT &442.7 $\pm$ 3.4 &1.00\\
2 &Normal Trace Compilation &1518.7 $\pm$ 7.2 &0.29\\
3 &Unrolling of Interp. Loop &737.6 $\pm$ 7.9 &0.60\\
4 &JIT, Full Optimizations &156.2 $\pm$ 3.8 &2.83\\
5 &Profile Overhead &515.0 $\pm$ 7.2 &0.86\\
\hline
\end{tabular}
\caption{Benchmark results of example interpreter computing the square of
\label{fig:bench1}
10000000}
\end{figure}

\begin{figure}[h]
\label{fig:bench-example}
{\small
\begin{verbatim}
def f(a):
    t = (1, 2, 3)
    i = 0
    while i < a:
        t = (t[1], t[2], t[0])
        i += t[0]
    return i
\end{verbatim}
}
\noindent
\begin{tabular}{|l|l|r|r|}
\hline
& &Time (s) &speedup\\
\hline
1 &Compiled to C, no JIT &23.44 $\pm$ 0.07 &1.00\\
2 &Compiled to C, with JIT &3.58 $\pm$ 0.05 &6.54\\
3 &CPython 2.5.2 &4.96 $\pm$ 0.05 &4.73\\
4 &CPython 2.5.2 + Psyco 1.6 &1.51 $\pm$ 0.05 &15.57\\\hline
\end{tabular}

@@

The results show that the tracing JIT speeds up the execution of this Python
function significantly, even outperforming CPython. To achieve this, the tracer
traces through the whole Python dispatching machinery, automatically inlining
the relevant fast paths. However, the manually tuned Psyco still performs a lot
better than our prototype.

@@

There have been some attempts to do \emph{dynamic partial evaluation}, which is
partial evaluation that defers partial evaluation completely to runtime
to make partial evaluation more useful for dynamic languages. This concept was
introduced by Sullivan \cite{sullivan_dynamic_2001} who implemented it for a
small dynamic language based on lambda-calculus. There is some work by one of
the authors to implement a dynamic partial evaluator for Prolog
\cite{carl_friedrich_bolz_automatic_2008}. There are also experiments within the
PyPy project to use dynamic partial evaluation for automatically generating JIT
compilers out of interpreters \cite{armin_rigo_jit_2007}. So far those have not been as
successful as we would like and it seems likely that they will be supplanted
with the work on tracing JITs described here.

@@

Furthermore both tracing and leaving machine code are very slow due to a
double interpretation overhead and we might need techniques for improving those.
Eventually we will need to apply the JIT to the various interpreters that are
written in RPython to evaluate how widely applicable the described techniques
are. Possible targets for such an evaluation would be the SPy-VM, a Smalltalk
implementation \cite{bolz_back_2008}; a Prolog interpreter or PyGirl, a Gameboy
emulator \cite{camillo_bruni_pygirl:_2009}; but also not immediately obvious
ones, like Python's regular expression engine.

>>>>

Let's look at how hints would need to be applied to the example interpreter
from Figure \ref{fig:tlr-basic}. Figure \ref{fig:tlr-full} shows the relevant
parts of the interpreter with hints applied.  One needs to instantiate
\texttt{JitDriver} by listing all the variables of the bytecode loop.
The variables are classified into two groups, ``green'' variables and ``red''
variables. The green variables are those that the tracing JIT should consider to
be part of the program counter of the language interpreter. In the case of the
example, the \texttt{pc} variable is obviously part of the program counter;
however, the \texttt{bytecode} variable is also counted as green, since the
\texttt{pc} variable is meaningless without the knowledge of which bytecode
string is currently being interpreted. All other variables are red (the fact
that red variables need to be listed explicitly too is an implementation
detail).

@@

For the small example the hints look like a lot of work. However, the number of
hints is essentially constant no matter how large the interpreter is, which
makes the extra work negligible for larger interpreters.

@@

The simple insight on how to improve the situation is that most of the operations
in the trace are actually concerned with manipulating the bytecode string and
the program counter. Those are stored in variables that are ``green'' (\ie they
are part of the position key).  This means that the tracer checks that those
variables have some fixed value at the beginning of the loop (they may well
change over the course of the loop, though). In the example of Figure
\ref{fig:trace-no-green-folding} the check would be that at the beginning of the
trace the \texttt{pc} variable is \texttt{4} and the \texttt{bytecode} variable
is the bytecode string corresponding to the square function. Therefore it is
possible to constant-fold computations on them away,
as long as the operations are side-effect free. Since strings are immutable in
RPython, it is possible to constant-fold the \texttt{strgetitem} operation. The
\texttt{int\_add} are additions of the green variable \texttt{pc} and a constant
number, so they can be folded away as well.

@@

From then on things proceed as described in Section \ref{sect:tracing}. The
tracing interpreter tries to find a loop in the user program, if it finds one it
will produce machine code for that loop and this machine code will be
immediately executed. The machine code is executed until a guard fails. Then the
execution should fall back to normal interpretation by the language interpreter.
This falling back is possibly a complex process, since the guard failure can
have occurred arbitrarily deep in a helper function of the language interpreter,
which would make it hard to rebuild the state of the language interpreter and
let it run from that point (\eg this would involve building a potentially deep
C stack). Instead the falling back is achieved by a special \emph{fallback
interpreter} which runs the language interpreter and the user program from the
point of the guard failure. The fallback interpreter is essentially a variant of
the tracing interpreter that does not keep a trace. The fallback interpreter
runs until execution reaches a safe point where it is easy to let the C version
of the language interpreter resume its operation.\footnote{This is the only
reason for the \texttt{jit\_merge\_point} hint.}  This means that the
fallback interpreter executes at most one bytecode operation of the language
interpreter and then falls back to the C version of the language interpreter.
After this, the whole process of profiling may start again.

Machine code production is done via a well-defined interface to an assembler
backend. This makes it possible to easily port the tracing JIT to various
architectures (including, we hope, to virtual machines such as the JVM where
our backend could generate JVM bytecode at runtime). At the moment the only
implemented backend is a 32-bit Intel-x86 backend.

@@

The first round of benchmarks (Figure \ref{fig:bench1}) are timings of the
example interpreter given in Figure \ref{fig:tlr-basic} computing
the square of 10000000\footnote{The result will overflow, but for smaller numbers the
running time is not long enough to sensibly measure it.}
using the bytecode of Figure \ref{fig:square}. The results for various
configurations are as follows:

@@

\textbf{Benchmark 3:} The tracing JIT is enabled and hints as in Figure
\ref{fig:tlr-full} are applied. This means that the interpreter loop is unrolled
so that it corresponds to the loop in the square function. However, constant folding of green
variables is disabled, therefore the resulting machine code corresponds to the
trace in Figure \ref{fig:trace-no-green-folding}. This alone brings no
improvement over the previous case.

@@

\begin{figure}
\noindent
\begin{tabular}{|l|l|r|r|}
\hline
& &Time (ms) &speedup\\
\hline
1 &Compiled to C, no JIT &442.7 $\pm$ 3.4 &1.00\\
2 &Normal Trace Compilation &1518.7 $\pm$ 7.2 &0.29\\
3 &Unrolling of Interp. Loop &737.6 $\pm$ 7.9 &0.60\\
4 &JIT, Full Optimizations &156.2 $\pm$ 3.8 &2.83\\
5 &Profile Overhead &515.0 $\pm$ 7.2 &0.86\\
\hline
\end{tabular}
\caption{Benchmark results of example interpreter computing the square of
\label{fig:bench1}
10000000}
\end{figure}

\begin{figure}[h]
\label{fig:bench-example}
{\small
\begin{verbatim}
def f(a):
    t = (1, 2, 3)
    i = 0
    while i < a:
        t = (t[1], t[2], t[0])
        i += t[0]
    return i
\end{verbatim}
}
\noindent
\begin{tabular}{|l|l|r|r|}
\hline
& &Time (s) &speedup\\
\hline
1 &PyPy compiled to C, no JIT &23.44 $\pm$ 0.07 &1.00\\
2 &PyPy comp'd to C, with JIT &3.58 $\pm$ 0.05 &6.54\\
3 &CPython 2.5.2 &4.96 $\pm$ 0.05 &4.73\\
4 &CPython 2.5.2 + Psyco 1.6 &1.51 $\pm$ 0.05 &15.57\\\hline
\end{tabular}

@@

The results show that the tracing JIT speeds up the execution of this Python
function significantly, even outperforming CPython. To achieve this, the tracer
traces through the whole Python dispatching machinery, automatically inlining
the relevant fast paths. However, the manually tuned Psyco still performs a lot
better than our prototype (although it is interesting to note that Psyco
improves the speed of CPython by only a factor 3.29 in this example, while
our tracing JIT improves PyPy by a factor 6.54).

@@

There have been some attempts to do \emph{dynamic partial evaluation}, which is
partial evaluation that defers partial evaluation completely to runtime
to make partial evaluation more useful for dynamic languages. This concept was
introduced by Sullivan \cite{sullivan_dynamic_2001} who implemented it for a
small dynamic language based on lambda-calculus. It is also related to
Psyco \cite{rigo_representation-based_2004}, a specializing JIT compiler for
Python.  There is some work by one of
the authors to implement a dynamic partial evaluator for Prolog
\cite{carl_friedrich_bolz_automatic_2008}. There are also experiments within the
PyPy project to use dynamic partial evaluation for automatically generating JIT
compilers out of interpreters \cite{armin_rigo_jit_2007}\cite{antocuni_2009}.
So far those have not been as
successful as we would like and it seems likely that they will be supplanted
with the work on tracing JITs described here.

@@

Furthermore both tracing and leaving machine code are very slow due to a
double interpretation overhead and we might need techniques for improving those.

Eventually we will need to apply the JIT to the various interpreters that are
written in RPython to evaluate how widely applicable the described techniques
are. Possible targets for such an evaluation would be the SPy-VM, a Smalltalk
implementation \cite{bolz_back_2008}; a Prolog interpreter; PyGirl, a Gameboy
emulator \cite{camillo_bruni_pygirl:_2009}; and also not immediately obvious
ones, like Python's regular expression engine.
